Monte Carlo Tree Search Applied to Co-operative Problems

### 2015-08-14 Report/ceec.tex
Final version started

%\author{Piers R. Williams, \and Joseph Walton-Rivers}

\title{MCTS Applied to Co-operative Problems}

@@

\newacronym{mcts}{MCTS}{Monte-Carlo Tree Search}
\newacronym{uct}{UCT}{Upper Confidence bound applied to Trees}
\newacronym{ggp}{GGP}{General Game Playing}
\newacronym{ga}{GA}{Genetic Algorithm}
\newacronym{mas}{MAS}{Multi-Agent Systems}
\newacronym{maga}{MAGA}{Macro Action Genetic Algorithm}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{fps}{FPS}{First-Person Shooter}
\newacronym{rts}{RTS}{Real-Time Strategy}
\newacronym{rpg}{RPG}{Role-Playing Game}

\begin{document}
\maketitle
\begin{abstract}
This paper highlights an experiment to see how standard MCTS handles simple co-operative problems with no prior or provided knowledge. These problems are formed from a simple grid world that has a set of goals, doors and buttons as well as walls that cannot be walked through. Two agents have to each reach every goal present on the map. For a door to be open, an Agent must be present on at least one of the buttons that is linked to it.

When laid out correctly, the world requires each agent to do certain things at certain times in order to achieve the goal. With no modification to allow communication between the two Agents - MCTS performs well and performs very 'purposefully' when given enough computational time.
\end{abstract}

\section{Introduction}
This research problem that this paper will asses is how \gls{ggp} agents perform when trying to solve a simple co-operative problem without co-operative abilities, with a focus on \gls{mcts}.

\gls{ggp} is the field of writing \gls{ai} agents that can play a multitude of games without being written specifically for each one individually \cite{genesereth2005general}. \gls{ggp} in real time video games is a popular competition \cite{perez2014}.

@@

\gls{mcts} \cite{browne2012survey} has been applied to a wealth of domains and is one of the primary algorithms in use for \gls{ggp} \cite{finnsson2008simulation}. Primary advantages are that, when given a sufficient forward model, \gls{mcts} doesn't require any strategic knowledge about the game itself in order to play.

Standard \gls{mcts} plays best when it is able to search far enough to locate states that provide a reward. \gls{mcts} tends to stumble when the time or search depth limit given to it isn't sufficient to allow it to locate any sequence of actions that provides a reward in order to differentiate the root's children \cite{perez2012monte}. When \gls{mcts} can't find any single state that contained a reward, all children of the node will contain identical values and the algorithm will be forced to make a random choice.

@@

\paragraph{\emph{SingleDoor}} This map is a basic map with two rooms, single door and a button either side.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{InitialState}
\caption{Single Door \protect\footnotemark}
\label{SingleDoor}
\end{figure}
\footnotetext{Displayed numbers are the ID numbers - \emph{Button} ID 0 will open \emph{Door} ID 0 but not \emph{Door} ID 1}
\paragraph{\emph{Pathfinding}} This map is the same as level1.txt but without the \emph{Doors} and \emph{Buttons}. This was to test the \emph{Agent's} ability to solve the simplest problem.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level1e}
\caption{Pathfinding}
\label{Pathfinding}
\end{figure}
\paragraph{\emph{SymmetricSingleDoor}} This map was a modification of the level1.txt - making the map more symmetric and the action parts closer together.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level2}
\caption{Symmetric Single Door}
\label{SymmetricSingleDoor}
\end{figure}
\paragraph{\emph{ExtendedSide}} This map was an extension of level1.txt with a second \emph{Goal} and some extra walls in the game. Spreading the action parts away from the door and buttons was the main purpose of this one.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level3}
\caption{Extended Side}
\label{ExtendedSide}
\end{figure}
\paragraph{\emph{SideBySide}} This map was a modification of level1.txt - putting a second \emph{Goal} in and making the map partly mirrored. Each \emph{Agent} began in a separate room to each other instead of together.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level4}
\caption{Side By Side}
\label{SideBySide}
\end{figure}
\paragraph{\emph{Airlock}} This map was a new map, designed to give wildly different roles to each AI - the top Agent would have to travel through two doors (Like an airlock) that the bottom agent had access to the linked \emph{Buttons}. Then the top \emph{Agent} could collect the reward and allow the bottom \emph{Agent} access to the \emph{Goal}.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level5}
\caption{Airlock}
\label{Airlock}
\end{figure}
\paragraph{\emph{Butterfly}} This map was a new map - with two rooms each with a \emph{Goal} and a door. This was designed to force each \emph{Agent} to be let into the \emph{Goal} room, and out of it again.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level6}
\caption{Butterfly}
\label{Butterfly}
\end{figure}

\subsubsection{\gls{ai} Controllers}
A set of \gls{ai} Controllers were created in order to attempt to solve the problem domain. None of the agents have the ability to communicate with another agent and each agent has a single controller.
\paragraph{Random}
The random controller simply chooses one of the possible 5 actions. This is one of the simplest to implement and run.
\begin{comment}
\paragraph{Keyboard}
The keyboard controller takes input from the keyboard and translates that into a single Action. Once an action is polled, it is reset to the No Op action unless the keyboard is used again. 
\end{comment}
\paragraph{\gls{mcts}}
The \gls{mcts} controller is a simple implementation of \gls{uct} - with a fixed number of rollouts, \gls{uct} tree search depth limit and rollout border. The rollout border is how far in total, the forward model will be allowed to progress before the game is evaluated. No knowledge about the game is provided and the assumption is made that the other player will play randomly. The fixed rollout border makes a significant improvement in iterations per decision (from 1-3 to 500-600 in 40ms \footnote{Intel Core i5-3570, 8GB RAM, Windows 7 Enterprise 64bit}). This greatly improves the ability of \gls{mcts} to make informed decisions. The score at the end of a rollout is taken from the game state - so if \gls{mcts} doesn't see any player reach a goal, all branches will be equal.

@@

\paragraph{Macro Action Genetic Algorithm}
The \gls{ga} algorithm was selected for its simplicity and had Macro Actions added in order to improve the forward search capability as well as the amount of computation time per decision available to it. The \gls{maga} used a population size of 10 and tournament selection of 3. Each candidate was a string of 15 actions - with each action performed 3 times in a row. This meant the \gls{maga} could "see" 45 ticks in the future. The design of the \gls{maga} algorithm was inspired by \textit{Perez et al} \cite{perez2013rolling}
\paragraph{Variable Macro Action Genetic Algorithm}
The Variable Macro Action \gls{ga} was created to try to solve the shortcomings of \gls{maga}. The primary premise was to allow the \gls{ga} to evolve the individual macro action lengths and the length of the overall sequence (number of macro actions). The main technique used was a 1 + 1 Evolutionary Strategy \cite{t.back2000evolutionary-co}. A 1 + 1 ES is a very simple \gls{ga} that maintains a single candidate, and in each iteration a mutation is made and if it is an improvement it is saved as the new best candidate. It is discarded if it is not an improvement. ES's are often much simpler than full gls{ga} algorithms to create, often require much less memory and due to performing less within each iteration is able to use more of the available time budget than a full \gls{ga}. The algorithm was bounded by a number of parameters that are shown in Table \ref{vmagaTable}.
\begin{table}[!t]
\begin{center}
\begin{tabular}{| l | l | l |}
\hline
\textbf{Param} & \textbf{Controls} & \textbf{Value} \\
\hline
minNum & Number of Macro Actions & 3 \\
MaxNum & Number of Macro Actions & 10 \\
minLength & Length of Macro Actions & 1 \\
maxLength & Length of Macro Actions & 5 \\
numChance & Chance of altering Number & 0.25 \\
lengthChance & Chance of altering each length & 0.8 \\
actionChance & Chance of altering each action & 0.75 \\
\hline
\end{tabular}
\caption{Table of parameters for the Variable Macro Action Genetic Algorithm}
\label{vmagaTable}
\end{center}
\end{table}
These values need extensive tuning to realise the full potential of the algorithm and is a source of possible future work.
\section{Results}

@@

Level 5 - shown in Figure \ref{Airlock} - poses a particular problem for many of the agents (see Figure \ref{avgScoreMap5}). The asymmetric nature of the level, and the delayed reward caused great difficulty for the Agents. Relying on two button presses and the other agent to get through the open doors - in order - greatly reduced performance. The top agent, \emph{High MCTS}, only managed an average score of $0.44$ compared to its total average of just over $0.8$.

\subsection{Discussion}
\subsubsection{MCTS}
\gls{mcts} performed very well in this problem domain.

@@

One possible reason for \gls{mcts} scoring so well in this problem domain is its use of statistics over hundreds and thousands of simulations to provide it with the ability to act in such a way that it handles all eventualities. Statistically,  in certain situations \gls{mcts} would only see rewards in the tree when it was situated on a \emph{Button}. This tended to cause the \gls{mcts} to travle towards \emph{Buttons}, increasing the likelihood that its own simulations would cause it to be situated on the button. Eventually, most other \gls{ai} \emph{Agents} would cross through the open door.
\subsubsection{GA}
The two \gls{ga} algorithms did not perform very well in this problem domain.

As seen in Figure \ref{avgScorePairAllMaps} and Figure \ref{avgTicksPairAllMaps}, when tasked with solving the problem domain with another identical agent, neither the \emph{GAController} or \emph{VariGA} performed well at all. The \emph{VariGA} only performed significantly better than either \emph{RandomController} or \emph{GAController} in \emph{level1E.txt}. In all other cases, \emph{VariGA} performed similarly to the standard \emph{GAController}. The ability to mutate the lengths of individual action sequences made the \emph{VariGA} a more flexible pathfinder that the \emph{GAController} but did not aid its ability to solve the co-operative problems present elsewhere in the experiment.

The \gls{ga} algorithms don't have the stored tree structure of \gls{mcts} with which to gain the statistical model for what happens when they pursue certain actions. This leads to a seemingly poor performance for the \gls{ga} despite \gls{ga} and \gls{mcts} typically performing equally well in other domains.
\subsubsection{Random}
The \emph{RandomController} performed poorly in this problem domain. On average, the \emph{RandomController} finished under the 2000 tick limit and scored less than half the maximum on average.
\section{Conclusions}
In this paper, we found that a strong MCTS player can solve simple cooperative problems without requiring communication between agents. We also found that a number of other \gls{ai} techniques experienced difficulty when the problem became cooperative. Further work is required in this area.

@@

\subsubsection{Optimisations}
Future work should also include some work to ensure that the \gls{ai} Agents have their parameters set correctly. The \emph{High MCTS} performed well - showing the power of having the correct parameters compared to the \emph{Low MCTS}. The \emph{VariGA} has 7 parameters that we simply didn't have time to correctly optimise for this experiment.

>>>>

\author{Piers R. Williams, \and Joseph Walton-Rivers}

\title{Monte Carlo Tree Search Applied to Co-operative Problems}

@@

\newacronym{mcts}{MCTS}{Monte Carlo Tree Search}
\newacronym{uct}{UCT}{Upper Confidence bound applied to Trees}
\newacronym{ggp}{GGP}{General Game Playing}
\newacronym{ga}{GA}{Genetic Algorithm}
\newacronym{mas}{MAS}{Multi-Agent Systems}
\newacronym{maga}{MAGA}{Macro Action Genetic Algorithm}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{fps}{FPS}{First-Person Shooter}
\newacronym{rts}{RTS}{Real-Time Strategy}
\newacronym{rpg}{RPG}{Role-Playing Game}

\begin{document}
\maketitle
\begin{abstract}
This paper highlights an experiment to see how standard Monte Carlo Tree Search handles simple co-operative problems with no prior or provided knowledge. These problems are formed from a simple grid world that has a set of goals, doors and buttons as well as walls that cannot be walked through. Two agents have to each reach every goal present on the map. For a door to be open, an Agent must be present on at least one of the buttons that is linked to it.

When laid out correctly, the world requires each agent to do certain things at certain times in order to achieve the goal. With no modification to allow communication between the two Agents - Monte Carlo Tress Search performs well and performs very 'purposefully' when given enough computational time.
\end{abstract}

\section{Introduction}
This research problem assessed in this paper is how \gls{ggp} agents perform when trying to solve a simple co-operative problem without co-operative abilities, with a focus on \gls{mcts}.

\gls{ggp} is the field of writing \gls{ai} agents that can play a multitude of games without being written specifically for each one individually \cite{genesereth2005general}. \gls{ggp} in real time video games has a popular competition \cite{perez2014} run frequently.

@@

\gls{mcts} \cite{browne2012survey} has been applied to a wealth of domains and is one of the primary algorithms in use for \gls{ggp} \cite{finnsson2008simulation}. Primary advantages are that, when given a sufficient forward model, \gls{mcts} does not require any strategic knowledge about the game itself in order to play.

Standard \gls{mcts} plays best when it is able to search far enough to locate states that provide a reward. \gls{mcts} tends to stumble when the time or search depth limit given to it is not sufficient to allow it to locate any sequence of actions that provides a reward in order to differentiate the root's children \cite{perez2012monte}. When \gls{mcts} can not find any single state that contained a reward, all children of the node will contain identical values and the algorithm will be forced to make a random choice.

@@

\paragraph{\emph{SingleDoor}} This map is a basic map with two rooms, single door and a button either side.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{InitialState}
\caption{Single Door \protect\footnotemark}
\label{SingleDoor}
\end{figure}
\footnotetext{Displayed numbers are the ID numbers - \emph{Button} ID 0 will open \emph{Door} ID 0 but not \emph{Door} ID 1}
\paragraph{\emph{Pathfinding}} This map is the same as \emph{SingleDoor} but without the \emph{Doors} and \emph{Buttons}. This was to test the \emph{Agent's} ability to solve the simplest problem.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level1e}
\caption{Pathfinding}
\label{Pathfinding}
\end{figure}
\paragraph{\emph{SymmetricSingleDoor}} This map was a modification of the \emph{SingleDoor} - making the map more symmetric and the action parts closer together.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level2}
\caption{Symmetric Single Door}
\label{SymmetricSingleDoor}
\end{figure}
\paragraph{\emph{ExtendedSide}} This map was an extension of \emph{SingleDoor} with a second \emph{Goal} and some extra walls in the game. Spreading the action parts away from the door and buttons was the main purpose of this one.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level3}
\caption{Extended Side}
\label{ExtendedSide}
\end{figure}
\paragraph{\emph{SideBySide}} This map was a modification of \emph{SingleDoor} - putting a second \emph{Goal} in and making the map partly mirrored. Each \emph{Agent} began in a separate room to each other instead of together.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level4}
\caption{Side By Side}
\label{SideBySide}
\end{figure}
\paragraph{\emph{Airlock}} This map was a new map, designed to give wildly different roles to each AI - the top Agent would have to travel through two doors (Like an airlock) that the bottom agent had access to the linked \emph{Buttons}. Then the top \emph{Agent} could collect the reward and allow the bottom \emph{Agent} access to the \emph{Goal}.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level5}
\caption{Airlock}
\label{Airlock}
\end{figure}
\paragraph{\emph{Butterfly}} This map was a new map - with two rooms each with a \emph{Goal} and a door. This was designed to force each \emph{Agent} to be let into the \emph{Goal} room, and out of it again.
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{level6}
\caption{Butterfly}
\label{Butterfly}
\end{figure}

\subsubsection{\gls{ai} Controllers}
A set of \gls{ai} Controllers were created in order to attempt to solve the problem domain. None of the agents have the ability to communicate with another agent and each agent has a single controller.
\paragraph{Random}
The random controller simply chooses one of the possible 5 actions. This is one of the simplest to implement and run.
\begin{comment}
\paragraph{Keyboard}
The keyboard controller takes input from the keyboard and translates that into a single Action. Once an action is polled, it is reset to the No Op action unless the keyboard is used again. 
\end{comment}
\paragraph{\gls{mcts}}
The \gls{mcts} controller is a simple implementation of \gls{uct} - with a fixed number of rollouts, \gls{uct} tree search depth limit and rollout border. The rollout border is how far in total, the forward model will be allowed to progress before the game is evaluated. No knowledge about the game is provided and the assumption is made that the other player will play randomly. The fixed rollout border makes a significant improvement in iterations per decision (from 1-3 to 500-600 in 40ms \footnote{Intel Core i5-3570, 8GB RAM, Windows 7 Enterprise 64bit}). This greatly improves the ability of \gls{mcts} to make informed decisions. The score at the end of a rollout is taken from the game state - so if \gls{mcts} does not see any player reach a goal, all branches will be equal.

@@

\paragraph{Macro Action Genetic Algorithm}
The \gls{ga} algorithm was selected for its simplicity and had Macro Actions added in order to improve the forward search capability as well as the amount of computation time per decision available to it. The \gls{maga} used a population size of 10 and tournament selection of 3. Each candidate was a string of 15 actions - with each action performed 3 times in a row. This meant the \gls{maga} could "see" 45 ticks in the future. The design of the \gls{maga} algorithm was inspired by \textit{Perez et al} \cite{perez2013rolling}
\paragraph{Variable Macro Action Genetic Algorithm}
The Variable Macro Action \gls{ga} was created to try to solve the shortcomings of \gls{maga}. The primary premise was to allow the \gls{ga} to evolve the individual macro action lengths and the length of the overall sequence (number of macro actions). The main technique used was a 1 + 1 Evolutionary Strategy \cite{t.back2000evolutionary-co}. A 1 + 1 ES is a very simple \gls{ga} that maintains a single candidate, and in each iteration a mutation is made and if it is an improvement it is saved as the new best candidate. It is discarded if it is not an improvement. ES's are often much simpler than full \gls{ga} algorithms to create, often require much less memory and due to performing less within each iteration is able to use more of the available time budget than a full \gls{ga}. The algorithm was bounded by a number of parameters that are shown in Table \ref{vmagaTable}.
\begin{table}[!t]
\begin{center}
\begin{tabular}{| l | l | l |}
\hline
\textbf{Param} & \textbf{Controls} & \textbf{Value} \\
\hline
minNum & Number of Macro Actions & 3 \\
MaxNum & Number of Macro Actions & 10 \\
minLength & Length of Macro Actions & 1 \\
maxLength & Length of Macro Actions & 5 \\
numChance & Chance of altering Number & 0.25 \\
lengthChance & Chance of altering each length & 0.8 \\
actionChance & Chance of altering each action & 0.75 \\
\hline
\end{tabular}
\caption{Table of parameters for the Variable Macro Action Genetic Algorithm}
\label{vmagaTable}
\end{center}
\end{table}
These values need extensive tuning to realise the full potential of the algorithm and is a source of possible future work.
\section{Results}

@@

Airlock - shown in Figure \ref{Airlock} - poses a particular problem for many of the agents (see Figure \ref{avgScoreMap5}). The asymmetric nature of the level, and the delayed reward caused great difficulty for the Agents. Relying on two button presses and the other agent to get through the open doors - in order - greatly reduced performance. The top agent, \emph{High MCTS}, only managed an average score of $0.44$ compared to its total average of just over $0.8$.

\section{Discussion}
\subsection{MCTS}
\gls{mcts} performed very well in this problem domain.

@@

One possible reason for \gls{mcts} scoring so well in this problem domain is its use of statistics over hundreds and thousands of simulations to provide it with the ability to act in such a way that it handles all eventualities. Statistically,  in certain situations \gls{mcts} would only see rewards in the tree when it was situated on a \emph{Button}. This tended to cause the \gls{mcts} to travle towards \emph{Buttons}, increasing the likelihood that its own simulations would cause it to be situated on the button. Eventually, most other \gls{ai} \emph{Agents} would cross through the open door.
\subsection{GA}
The two \gls{ga} algorithms did not perform very well in this problem domain.

As seen in Figure \ref{avgScorePairAllMaps} and Figure \ref{avgTicksPairAllMaps}, when tasked with solving the problem domain with another identical agent, neither the \emph{GAController} or \emph{VariGA} performed well at all. The \emph{VariGA} only performed significantly better than either \emph{RandomController} or \emph{GAController} in \emph{Pathfinding}. In all other cases, \emph{VariGA} performed similarly to the standard \emph{GAController}. The ability to mutate the lengths of individual action sequences made the \emph{VariGA} a more flexible pathfinder that the \emph{GAController} but did not aid its ability to solve the co-operative problems present elsewhere in the experiment.

The \gls{ga} algorithms do not have the stored tree structure of \gls{mcts} with which to gain the statistical model for what happens when they pursue certain actions. This leads to a seemingly poor performance for the \gls{ga} despite \gls{ga} and \gls{mcts} typically performing equally well in other domains.
\subsection{Random}
The \emph{RandomController} performed poorly in this problem domain. On average, the \emph{RandomController} finished under the 2000 tick limit and scored less than half the maximum on average.
\section{Conclusions}
In this paper, we found that a strong MCTS player can solve simple cooperative problems without requiring communication between agents. We also found that a number of other \gls{ai} techniques experienced difficulty when the problem became cooperative. Further work is required in this area.

@@

\subsubsection{Optimisations}
Future work should also include some work to ensure that the \gls{ai} Agents have their parameters set correctly. The \emph{High MCTS} performed well - showing the power of having the correct parameters compared to the \emph{Low MCTS}. The \emph{VariGA} has 7 parameters that we simply did not have time to correctly optimise for this experiment.
